---
title: "Lecture 3: ChemmineR"
---


```{r style, echo = FALSE, results = 'asis'}
options(width=100, max.print=1000)
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
    library(ChemmineR)
    library(fmcsR)
    library(ggplot2)
	 #library(ChemmineOB)
})
```

Note: the most recent version of this tutorial can be found <a href="https://htmlpreview.github.io/?https://github.com/girke-lab/ChemmineR/blob/master/vignettes/ChemmineR.html">here</a> and a short overview slide show [here](http://faculty.ucr.edu/~tgirke/HTML_Presentations/Manuals/Workshop_Dec_5_8_2014/Rcheminfo/Cheminfo.pdf).


`ChemmineR` is a cheminformatics package for analyzing
drug-like small molecule data in R. Its latest version contains
functions for efficient processing of large numbers of small molecules,
physicochemical/structural property predictions, structural similarity
searching, classification and clustering of compound libraries with a
wide spectrum of algorithms.


## Installation
------------

The R software for running ChemmineR can be downloaded from CRAN
(<http://cran.at.r-project.org/>). The ChemmineR package can be
installed from R with:

```{r eval=FALSE}
 if (!requireNamespace("BiocManager", quietly=TRUE))
     install.packages("BiocManager")
 BiocManager::install("ChemmineR")
```

Loading the Package and Documentation
-------------------------------------


```{r eval=TRUE, tidy=FALSE}
 library("ChemmineR") # Loads the package
```

```{r eval=FALSE, tidy=FALSE}
 library(help="ChemmineR") # Lists all functions and classes
 vignette("ChemmineR") # Opens this PDF manual from R
```


## Five Minute Tutorial
--------------------

The following code gives an overview of the most important
functionalities provided by `ChemmineR`. Copy and paste
of the commands into the R console will demonstrate their utilities.

Create Instances of `SDFset` class:
```{r eval=TRUE, tidy=FALSE}
 data(sdfsample)
 sdfset <- sdfsample
 sdfset # Returns summary of SDFset
 sdfset[1:4] # Subsetting of object

 sdfset[[1]] # Returns summarized content of one SDF


```{r eval=FALSE, tidy=FALSE}
 view(sdfset[1:4]) # Returns summarized content of many SDFs, not printed here
 as(sdfset[1:4], "list") # Returns complete content of many SDFs, not printed here

```


An `SDFset` is created during the import of an SD file:

```{r eval=FALSE, tidy=FALSE}
 sdfset <- read.SDFset("http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/Samples/sdfsample.sdf")
```


Miscellaneous accessor methods for `SDFset` container:

```{r eval=FALSE, tidy=FALSE}
 header(sdfset[1:4]) # Not printed here
```

```{r eval=TRUE, tidy=FALSE}
 header(sdfset[[1]])
```

```{r eval=FALSE, tidy=FALSE}
 atomblock(sdfset[1:4]) # Not printed here
```

```{r eval=TRUE, tidy=FALSE}
atomblock(sdfset[[1]])[1:4,]
```

```{r eval=FALSE, tidy=FALSE}
bondblock(sdfset[1:4]) # Not printed here
```

```{r eval=TRUE, tidy=FALSE}
 bondblock(sdfset[[1]])[1:4,]
```
```{r eval=FALSE, tidy=FALSE}
 datablock(sdfset[1:4]) # Not printed here
```

```{r eval=TRUE, tidy=FALSE}
 datablock(sdfset[[1]])[1:4]
```


Assigning compound IDs and keeping them unique:
```{r eval=TRUE, tidy=FALSE}
 cid(sdfset)[1:4] # Returns IDs from SDFset object
 sdfid(sdfset)[1:4] # Returns IDs from SD file header block
 unique_ids <- makeUnique(sdfid(sdfset))
 cid(sdfset) <- unique_ids
```


Converting the data blocks in an `SDFset` to a matrix:

```{r eval=TRUE, tidy=FALSE}
 blockmatrix <- datablock2ma(datablocklist=datablock(sdfset)) # Converts data block to matrix
 numchar <- splitNumChar(blockmatrix=blockmatrix) # Splits to numeric and character matrix
 numchar[[1]][1:2,1:2] # Slice of numeric matrix
 numchar[[2]][1:2,10:11] # Slice of character matrix
```


Compute atom frequency matrix, molecular weight and formula:

```{r eval=TRUE, tidy=FALSE}
 propma <- data.frame(MF=MF(sdfset), MW=MW(sdfset), atomcountMA(sdfset))
 propma[1:4, ]
```


Assign matrix data to data block:
```{r eval=TRUE, tidy=FALSE}
 datablock(sdfset) <- propma
 datablock(sdfset[1])
```


String searching in `SDFset`:
```{r eval=FALSE, tidy=FALSE}
 grepSDFset("650001", sdfset, field="datablock", mode="subset") # Returns summary view of matches. Not printed here.
```

```{r eval=TRUE, tidy=FALSE}
 grepSDFset("650001", sdfset, field="datablock", mode="index")
```


Export SDFset to SD file:
```{r eval=FALSE, tidy=FALSE}
 write.SDF(sdfset[1:4], file="sub.sdf", sig=TRUE)
```


Plot molecule structure of one or many SDFs:
```{r plotstruct, eval=TRUE, tidy=FALSE}
 plot(sdfset[1:4], print=FALSE) # Plots structures to R graphics device
```



```{r eval=FALSE, tidy=FALSE}
 sdf.visualize(sdfset[1:4]) # Compound viewing in web browser
```

![Figure: Visualization webpage created by calling
`sdf.visualize`.](visualizescreenshot-small.png )

Structure similarity searching and clustering:
```{r eval=FALSE, tidy=FALSE}
 apset <- sdf2ap(sdfset) # Generate atom pair descriptor database for searching
```

```{r eval=TRUE, tidy=FALSE}
 data(apset) # Load sample apset data provided by library.
 cmp.search(apset, apset[1], type=3, cutoff = 0.3, quiet=TRUE) # Search apset database with single compound.

 cmp.cluster(db=apset, cutoff = c(0.65, 0.5), quiet=TRUE)[1:4,] # Binning clustering using variable similarity cutoffs.
```


### Molecular Structure Data
------------------------

Classes

-   `SDFstr`: intermediate string class to facilitate SD
    file import; not important for end user

-   `SDF`: container for single molecule imported from an
    SD file

-   `SDFset`: container for many SDF objects; most
    important structure container for end user

-   `SMI`: container for a single SMILES string

-   `SMIset`: container for many SMILES strings

Functions/Methods (mainly for `SDFset` container,
`SMIset` should be coerced with
`smiles2sd` to `SDFset`)

-   Accessor methods for `SDF/SDFset`

    -   Object slots: `cid`, `header`, `atomblock`, `bondblock`,
        `datablock` (`sdfid`, `datablocktag`)

    -   Summary of `SDFset`: `view`

    -   Matrix conversion of data block: `datablock2ma`,
        `splitNumChar`

    -   String search in SDFset: `grepSDFset`

-   Coerce one class to another

    -   Standard syntax `as(..., "...")` works in most
        cases. For details see R help with
       `?"SDFset-class"`.

-   Utilities

    -   Atom frequencies: `atomcountMA`, `atomcount`

    -   Molecular weight: `MW`

    -   Molecular formula: `MF`

    -   ...

-   Compound structure depictions

    -   R graphics device: `plot`, `plotStruc`

    -   Online: `cmp.visualize`

<div align="right">[Back to Table of Contents]()</div>

Structure Descriptor Data
-------------------------

Classes

-   `AP`: container for atom pair descriptors of a single
    molecule

-   `APset`: container for many AP objects; most
    important structure descriptor container for end user

-   `FP`: container for fingerprint of a single molecule

-   `FPset`: container for fingerprints of many
    molecules, most important structure descriptor container for end
    user

Functions/Methods

-   Create `AP/APset` instances

    -   From `SDFset`: `sdf2ap`

    -   From SD file: `cmp.parse`

    -   Summary of `AP/APset`: `view`,
        `db.explain`

-   Accessor methods for AP/APset

    -   Object slots: `ap`, `cid`

-   Coerce one class to another

    -   Standard syntax `as(..., "...")` works in most
        cases. For details see R help with
        `?"APset-class"`.

-   Structure Similarity comparisons and Searching

    -   Compute pairwise similarities : `cmp.similarity`,
        `fpSim`

    -   Search APset database: `cmp.search`, `fpSim`

-   AP-based Structure Similarity Clustering

    -   Single-linkage binning clustering: `cmp.cluster`

    -   Visualize clustering result with MDS:
        `cluster.visualize`

    -   Size distribution of clusters: `cluster.sizestat`
-   Folding
    - Fold a descriptor with `fold`
	 - Query the number of times a descriptor has been folded:
		`foldCount`
	 - Query the number of bits in a descriptor: `numBits`

<div align="right">[Back to Table of Contents]()</div>

Import of Compounds
===================

SDF Import
----------

The following gives an overview of the most important import/export
functionalities for small molecules provided by
`ChemmineR`. The given example creates an instance of the
`SDFset` class using as sample data set the first 100
compounds from this PubChem SD file (SDF):
Compound\_00650001\_00675000.sdf.gz
(<ftp://ftp.ncbi.nih.gov/pubchem/Compound/CURRENT-Full/SDF/>).

SDFs can be imported with the `read.SDFset` function:

```{r eval=FALSE, tidy=FALSE}
 sdfset <- read.SDFset("http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/Samples/sdfsample.sdf")
```

```{r eval=TRUE, tidy=FALSE}
 data(sdfsample) # Loads the same SDFset provided by the library
 sdfset <- sdfsample
 valid <- validSDF(sdfset) # Identifies invalid SDFs in SDFset objects
 sdfset <- sdfset[valid] # Removes invalid SDFs, if there are any
```


Import SD file into `SDFstr` container:
```{r eval=FALSE, tidy=FALSE}
 sdfstr <- read.SDFstr("http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/Samples/sdfsample.sdf")
```
Create
`SDFset` from `SDFstr` class:

```{r eval=TRUE, tidy=FALSE}
 sdfstr <- as(sdfset, "SDFstr")
 sdfstr
 as(sdfstr, "SDFset")
```

<div align="right">[Back to Table of Contents]()</div>

SMILES Import
-------------

The `read.SMIset` function imports one or many molecules
from a SMILES file and stores them in a `SMIset`
container. The input file is expected to contain one SMILES string per
row with tab-separated compound identifiers at the end of each line. The
compound identifiers are optional.

Create sample SMILES file and then import it:
```{r eval=FALSE, tidy=FALSE}
 data(smisample); smiset <- smisample
 write.SMI(smiset[1:4], file="sub.smi")
 smiset <- read.SMIset("sub.smi")
```


Inspect content of `SMIset`:
```{r eval=TRUE, tidy=FALSE}
 data(smisample) # Loads the same SMIset provided by the library
 smiset <- smisample
 smiset
 view(smiset[1:2])
```


Accessor functions:
```{r eval=TRUE, tidy=FALSE}
 cid(smiset[1:4])
 smi <- as.character(smiset[1:2])
```


Create `SMIset` from named character vector:

```{r eval=TRUE, tidy=FALSE}
 as(smi, "SMIset")
```

<div align="right">[Back to Table of Contents]()</div>

Export of Compounds
===================

SDF Export
----------

Write objects of classes `SDFset/SDFstr/SDF` to SD file:

```{r eval=FALSE, tidy=FALSE}
 write.SDF(sdfset[1:4], file="sub.sdf")
```


Writing customized `SDFset` to file containing
`ChemmineR` signature, IDs from `SDFset`
and no data block:
```{r eval=FALSE, tidy=FALSE}
 write.SDF(sdfset[1:4], file="sub.sdf", sig=TRUE, cid=TRUE, db=NULL)
```


Example for injecting a custom matrix/data frame into the data block of
an `SDFset` and then writing it to an SD file:

```{r eval=FALSE, tidy=FALSE}
 props <- data.frame(MF=MF(sdfset), MW=MW(sdfset), atomcountMA(sdfset))
 datablock(sdfset) <- props
 write.SDF(sdfset[1:4], file="sub.sdf", sig=TRUE, cid=TRUE)
```


Indirect export via `SDFstr` object:
```{r eval=FALSE, tidy=FALSE}
 sdf2str(sdf=sdfset[[1]], sig=TRUE, cid=TRUE) # Uses default components
 sdf2str(sdf=sdfset[[1]], head=letters[1:4], db=NULL) # Uses custom components for header and data block
```


Write `SDF`, `SDFset` or
`SDFstr` classes to file:
```{r eval=FALSE, tidy=FALSE}
 write.SDF(sdfset[1:4], file="sub.sdf", sig=TRUE, cid=TRUE, db=NULL)
 write.SDF(sdfstr[1:4], file="sub.sdf")
 cat(unlist(as(sdfstr[1:4], "list")), file="sub.sdf", sep="")
```

<div align="right">[Back to Table of Contents]()</div>

SMILES Export
-------------

Write objects of class `SMIset` to SMILES file with and
without compound identifiers:
```{r eval=FALSE, tidy=FALSE}
 data(smisample); smiset <- smisample # Sample data set

 write.SMI(smiset[1:4], file="sub.smi", cid=TRUE) write.SMI(smiset[1:4], file="sub.smi", cid=FALSE)
```

<div align="right">[Back to Table of Contents]()</div>

Format Interconversions
=======================

The `sdf2smiles` and `smiles2sdf`
functions provide format interconversion between SMILES strings
(Simplified Molecular Input Line Entry Specification) and
`SDFset` containers.

Convert an `SDFset` container to a SMILES
`character` string:


```{r sdf2smiles, eval=FALSE, tidy=FALSE}
 data(sdfsample);
 sdfset <- sdfsample[1]
 smiles <- sdf2smiles(sdfset)
 smiles
```


Convert a SMILES `character` string to an
`SDFset` container:


```{r smiles2sdf, eval=FALSE, tidy=FALSE}
 sdf <- smiles2sdf("CC(=O)OC1=CC=CC=C1C(=O)O")
 view(sdf)
```


When the `ChemineOB` package is installed these
conversions are performed with the OpenBabel Open Source Chemistry
Toolbox. Otherwise the functions will fall back to using the ChemMine
Tools web service for this operation. The latter will require internet
connectivity and is limited to only the first compound given.
`ChemmineOB` provides access to the compound format
conversion functions of OpenBabel. Currently, over 160 formats are
supported by OpenBabel. The functions `convertFormat` and
`convertFormatFile` can be used to convert files or
strings between any two formats supported by OpenBabel. For example, to
convert a SMILES string to an SDF string, one can use the
`convertFormat` function.


```{r eval=FALSE, tidy=FALSE}
 sdfStr <- convertFormat("SMI","SDF","CC(=O)OC1=CC=CC=C1C(=O)O_name")
```


This will return the given compound as an SDF formatted string. 2D
coordinates are also computed and included in the resulting SDF string.

To convert a file with compounds encoded in one format to another
format, the `convertFormatFile` function can be used
instead.
```{r eval=FALSE, tidy=FALSE}
 convertFormatFile("SMI","SDF","test.smiles","test.sdf")
```


To see the whole list of file formats supported by OpenBabel, one can
run from the command-line "obabel -L formats".

<div align="right">[Back to Table of Contents]()</div>

Splitting SD Files
==================

The following `write.SDFsplit` function allows to split
SD Files into any number of smaller SD Files. This can become important
when working with very big SD Files. Users should note that this
function can output many files, thus one should run it in a dedicated
directory!

Create sample SD File with 100 molecules:
```{r eval=FALSE, tidy=FALSE}
 write.SDF(sdfset, "test.sdf")
```


Read in sample SD File. Note: reading file into SDFstr is much faster
than into SDFset:
```{r eval=FALSE, tidy=FALSE}
 sdfstr <- read.SDFstr("test.sdf")
```


Run export on `SDFstr` object:
```{r eval=FALSE, tidy=FALSE}
 write.SDFsplit(x=sdfstr, filetag="myfile", nmol=10) # 'nmol' defines the number of molecules to write to each file
```


Run export on `SDFset` object:
```{r eval=FALSE, tidy=FALSE}
 write.SDFsplit(x=sdfset, filetag="myfile", nmol=10)
```


Searching PubChem
=================

Get Compounds from PubChem by Id
--------------------------------

The function `getIds` accepts one or more numeric PubChem
compound ids and downloads the corresponding compounds from PubChem
Power User Gateway (PUG) returning results in an `SDFset`
container. The ChemMine Tools web service is used as an intermediate, to
translate queries from plain HTTP POST to a PUG SOAP query.

Fetch 2 compounds from PubChem:


```{r getIds, eval=FALSE, tidy=FALSE}
 compounds <- getIds(c(111,123))
 compounds
```

<div align="right">[Back to Table of Contents]()</div>

Search a SMILES Query in PubChem
--------------------------------

The function `searchString` accepts one SMILES string
(Simplified Molecular Input Line Entry Specification) and performs a
\>0.95 similarity PubChem fingerprint search, returning the hits in an
`SDFset` container. The ChemMine Tools web service is
used as an intermediate, to translate queries from plain HTTP POST to a
PubChem Power User Gateway (PUG) query.

Search a SMILES string on PubChem:


```{r searchString, eval=FALSE, tidy=FALSE}
 compounds <- searchString("CC(=O)OC1=CC=CC=C1C(=O)O") compounds
```

<div align="right">[Back to Table of Contents]()</div>

Search an SDF Query in PubChem
------------------------------

The function `searchSim` performs a PubChem similarity
search just like `searchString`, but accepts a query in
an `SDFset` container. If the query contains more than
one compound, only the first is searched.

Search an `SDFset` container on PubChem:


```{r searchSim, eval=FALSE, tidy=FALSE}
 data(sdfsample);
 sdfset <- sdfsample[1]
 compounds <- searchSim(sdfset)
 compounds
```

```{r knit_exit, include=F, echo=F}
knit_exit()
```
