---
title: "YUM 101 - Lectures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo =FALSE}

# devtools::install_github('yihui/xfun')

# a single RmD file
# xfun::embed_file('example.Rmd')
```


<br>


## History of Pizza
Wed, Sep 25  

Slides: `r xfun::embed_file('slides.pptx')`  
Readings: 

* [History of Pizza](https://en.wikipedia.org/wiki/History_of_pizza)

<br>

## Woodfired Thin Crust Pizza
Wed, Oct 2  

Slides: `r xfun::embed_file('slides.pptx')`  
Readings: 

* [History of Pizza](https://en.wikipedia.org/wiki/History_of_pizza)  

<br>


## Fusion Pizza and Creative Renditions
Wed, Oct 9 

Slides: `r xfun::embed_file('slides.pptx')`  
Readings: 

* [Article 1](https://en.wikipedia.org/wiki/History_of_pizza)  
 
<br>


## Intro to Pastries
Wed, Oct 16

Slides: `r xfun::embed_file('slides.pptx')`  
Readings: 

* [Article 2](https://en.wikipedia.org/wiki/History_of_pizza)  
 
<br>


## Danishes
Wed, Oct 23


Slides: `r xfun::embed_file('slides.pptx')`  
Readings: 

* [Article 3](https://en.wikipedia.org/wiki/History_of_pizza)  
 
<br>

## Icings, Frosing, & Glazes
Wed, Oct 30


Slides: `r xfun::embed_file('slides.pptx')`
Readings: 

* [Article 4](https://en.wikipedia.org/wiki/History_of_pizza)  
 
 
<br>


## Cakes, Cupcakes, and Rolls
Wed,  Nov 6

Readings: 

* [Article 5](https://en.wikipedia.org/wiki/History_of_pizza)  
 
 
<br>


## Cookies
Wed,  Nov 13

Readings: 

* [History of Cookies](https://en.wikipedia.org/wiki/Cookie)  
 
 
<br>


## Pies
Wed, Nov 20 

Slides: `r xfun::embed_file('slides.pptx')`  
Readings: 

* [Types of Pies](https://en.wikipedia.org/wiki/List_of_pies,_tarts_and_flans)  
 


