---
title:        "Lecture 3: R programming II"
# subtitle:   "Based on: https://github.com/mattiaguerini/slides-intro-to-R"
# author:     "Mariam R. Rizkallah"
# date:       "2021"
# output:     beamer_presentation
# theme:      boxes
# colortheme: dolphin
# fonttheme:  serif
# toc:        TRUE
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

# Basic programming

## Variables
In programming, a variable denotes an object (i.e. a variable is a name or a label for something).
\scriptsize
```{r variables, include=T, echo=T}
x <- 1
my_function <- function(x){x*2+2}
```
\normalsize
Notice that the argument x of the function is different from the x previously defined. The second is only local to the function and always required to be specified.

Try to compute `r my_function(1)` or `r my_function(9)`.

## Control Flows (I)
Also known as an if/else statement. It relates to ways in which you can adapt your code to different circumstances.

Based on a condition being TRUE, your program will do one thing, as opposed to another thing.

In R, the if/else syntax has the following structure:
\scriptsize
```{r controlflow_prepare, include=F, echo=F}
condition <- T
do_something <- "do something"
do_something_different <- "do something different"
```
\normalsize

\scriptsize
```{r controlflow, include=T, echo=T}
if (condition == TRUE) {
  do_something
} else {
  do_something_different
}
```
\normalsize

## Control Flows (II) - Example
\scriptsize
```{r controlflow2, include=T, echo=T}
x <- 1
y <- 3
if (x>y) {
  print("x is larger than y")
  z <- x*y
  print(paste0("z is equal to ", z))
} else {
  print("x is smaller or equal than y")
  z <- x*y - 1
  print(paste0("z is equal to ", z))
}
```
\normalsize

## Control Flows (III) - Example with more conditions
\scriptsize
```{r controlflow3, include=T, echo=T}
x <- 3
y <- 3
if (x>y) {
  print("x is larger than y")
  z <- x*y + 1
  print(paste0("z is equal to ", z))
} else if (x==y) {
  print("x is equal than y")
  z <- x*y
  print(paste0("z is equal to ", z))
} else {
  print("x is smaller than y")
  z <- x*y - 1
  print(paste0("z is equal to ", z))
}
```
\normalsize

## Loops (I)
As the name suggests, in a loop the program repeats a set of instructions many times, until some condition tells it to stop.

A very powerful, yet simple, construction is that the program can count how many steps it has done already - which may be important to know for many algorithms.

The syntax of a `for` loop is the following:
\scriptsize
```{r loops, include=T, echo=T}
for (i in 1:10){
  # does not have to be 1:10!
  # loop body: gets executed each time
  # the value of i changes with each iteration
}
```
\normalsize

## Loops (II) - Example
Produce a loop that displays the double of the loop round.
\scriptsize
```{r loops2, include=T, echo=T}
for (i in 1:5){
  y <- i*2
  print(y)
}
```
\normalsize

## Loops (III) - Example with more loops
You can even have loops into other loops.

These can be useful for exploring combinations of events:
\scriptsize
```{r loops3, include=T, echo=T}
quantity <- c(2,3)
fruits <- c("mangos", "apples", "bananas")

for (i in quantity){ # first nest: for each i
  for (j in fruits){ # second nest: for each j
    print(paste("Can I get",i,j,"please?"))
    }
}
```
\normalsize

## Functions (I)
So far we have been using functions, but haven’t actually discussed some of their details.

A function is a set of instructions that R executes for us, much like those collected in a script file.

The good thing is that functions are much more flexible than scripts, since they can depend on input arguments, which change the way the function behaves.

## Functions (II)
Here is how to define a function in general:
\scriptsize
```{r functions, include=T, echo=T}
function_name <- function(arg1 ,arg2=default_value){
  # function body
  # you do stuff with arg1 and arg2
  # you can have any number of arguments, with or without defaults
  # any valid `R` commands can be included here
  # the last line is returned
}
```
\normalsize

## Function (III) - Example
\scriptsize
```{r functions2, include=T, echo=T}
hello <- function(your_name = "Lord Vader"){
  paste("You R most welcome,", your_name)
  # we could also write:
  # return(paste("You R most welcome,",your_name))
}
# we call the function by typing it's name with round brackets
```

```{r functions3, include=T, echo=T}
hello()
hello("Mariam")
```
\normalsize

# Scripts
## Scripts
\scriptsize
```{r script1, include=T, echo=T, eval=F}
source(file = "sample_script.R")
```

```{r script2, include=T, echo=T}
# A comment: this is a sample script.
y <- c(12,15,28,17,18)
x <- c(22,39,50,25,18)
mean(y)
mean(x)
```
\normalsize


## Challenge
* Script and function (and parameters)
```{r script3, include=T, echo=T, eval=F}
Rscript myname.R Delta
```

```{r knit_exit, include=F, echo=F}
knit_exit()
```
